---
title: "Exploration"
output: html_document
date: "2025-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Init
```{r}
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(predictmeans)

library(lme4)
library(predictmeans)



```
# Test

```{r}
# setwd("C:/Users/huste/Documents/Github/businessAnalytics/Correlated Data/Assignment 1")
setwd("/home/husted42/Documents/github/businessAnalytics/Correlated Data/Assignment 1")

getwd()

list.files()

TV <- read.table("TV2.txt", header = 1)
TV$Assessor <- as.factor(TV$Assessor)
TV$Repeat   <- as.factor(TV$Repeat)
TV$TVset    <- as.factor(TV$TVset)
TV$Picture    <- as.factor(TV$Picture)


```

## Data
### Means
 - Assessor indicates the assessor (1-8), 
 - TVset the television setting (TV1,TV2,TV3),
 - Picture the picture (1-3), 
 - Repeat the replicate (0,1), 
 - (Target) Coloursaturation denotes the colour saturation score given by the assessor. 

```{r}
TV

vars <- c("TVset", "Assessor", "Repeat", "Picture")

par(mfrow = c(2, 2))

for (v in vars) {
  formula <- as.formula(paste("Coloursaturation ~", v))
  
  boxplot(formula, data = TV,
          main = paste("Colour Saturation by", v),
          xlab = v,
          ylab = "Colour Saturation",
          col = "lightblue")

          means <- tapply(TV$Coloursaturation, TV$TVset, mean)
          points(1:length(means), means, pch = 19, col = "black")
  
  }

par(mfrow = c(1, 1))

```
Black dot is mean
Black bar is the median 
Dotted lines shows the confidence interval
Blue box - Interquartile Range (IQR) (Shows the middle 50% of your data.)

We see a difference in mean and meadian in TVset for TV2. It's interquartile range doesn't overlap TV1 and TV3
Also low variance for Assessor 2 and 7

### Interactions
```{r}
# All pairwise interaction plots among these factors
vars <- c("TVset", "Repeat", "Picture")

op <- par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

pairs <- combn(vars, 2, simplify = FALSE)
for (p in pairs) {
  xfac <- p[[1]]
  trc  <- p[[2]]
  interaction.plot(
    x.factor   = TV[[xfac]],
    trace.factor = TV[[trc]],
    response   = TV$Coloursaturation,
    fun        = mean,
    type       = "b",
    pch        = 19,
    xlab       = xfac,
    ylab       = "Colour Saturation (mean)",
    legend     = TRUE,
    main       = paste("Interaction:", xfac, "×", trc)
  )
} 
```

In TVset:Repeat the lines follows each other meaning that it's unlikly there is an interaction. TVset doesn't seem to have an impact on Repeat and vice versa. The same follows for TVset:Picture. 

For repeat:picture it may look there is some interaction because the patterns doesn't follow each other as closely, but the interval between highest and lowest value are very low, which indicate that our model won't improve much by including it.


## Determining the model
### Interactions

```{r}
model_0 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                TVset:Picture + 
                Repeat:Picture +
                (1 | Assessor),
              
              data = TV
  )


model_1 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                TVset:Picture + 
                (1 | Assessor),
              data = TV
  )

model_2 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                (1 | Assessor),
              data = TV
  )

model_3 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                (1 | Assessor),
              data = TV
  )

a1 <- anova(model_0, model_1, refit = FALSE, type = 3)
a2 <- anova(model_1, model_2, refit = FALSE, type = 3)
a3 <- anova(model_2, model_3, refit = FALSE, type = 3)

cat("#################### Anova 1 - Repeat:Pictur ####################\n")
print(a1)
cat("p-value =", a1$`Pr(>Chisq)`[2], "\n\n")

cat("#################### Anova 2 - TVset:Picture ####################\n")
print(a2)
cat("p-value =", a2$`Pr(>Chisq)`[2], "\n\n")

cat("#################### Anova 3 - TVset:Repeat####################\n")
print(a3)
cat("p-value =", a3$`Pr(>Chisq)`[2], "\n\n")



```
We fail to reject the null hypothesis $ H_0 : \beta_{Repeat:Picture} = 0 $ on a 95% confidence interval with a p-value of 1. This indicates that there is no statistically significant interaction between Repeat and Picture. This holds for all three interactions on with p-values 0.41, 0.13, 0.30.

```{r}
library(lme4)

full_formula <- Coloursaturation ~ 
  TVset + Repeat + Picture +
  TVset:Repeat + TVset:Picture + Repeat:Picture +
  (1 | Assessor)

terms_to_drop <- c("Repeat:Picture", "TVset:Picture", "TVset:Repeat")

models <- vector("list", length(terms_to_drop) + 1)
names(models) <- c("full", paste0("drop_", terms_to_drop))
models[[1]] <- lmer(full_formula, data = TV)

for (i in seq_along(terms_to_drop)) {
  models[[i + 1]] <- update(models[[i]], paste(". ~ . -", terms_to_drop[i]))
}

comparisons <- vector("list", length(terms_to_drop))
names(comparisons) <- paste0("Test: drop ", terms_to_drop)

for (i in seq_along(terms_to_drop)) {
  comparisons[[i]] <- anova(models[[i]], models[[i + 1]], refit = FALSE, type = 3)
}

for (i in seq_along(comparisons)) {
  term <- terms_to_drop[i]
  a <- comparisons[[i]]
  pval <- a$`Pr(>Chisq)`[nrow(a)]
  cat("\n#################### Test -", term, "####################\n")
  print(a)
  cat(sprintf("p-value = %.4g\n", pval))
}
```


### Random effects
```{r}

model_effect <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                (1 | Assessor),
                data = TV
  )

ranova(model_effect)

```

The p-values is very small indicating systematically difference in how Assessors rate color saturation, so we keep the random effect. 


### Fixed effects
```{r}

anova(model_effect, type = 3)

```
We reject the null hypothesis that $ \mu_{TVset} = 0  $ and $ \beta_{picture} = 0  $ on a significance level of 5%. 
We drop Repeat since it doesn't improve the model.

### Final model
```{r}
model <- lmer(
  Coloursaturation ~ TVset + Picture + (1 | Assessor),
  data = TV
)

summary(model)
```
The mean of our reference TVset = TV1 and Picture = Picture1 is 7.4 and there is a higly significant increase when switching to TV setting TV2 of 1.97 on average. Also the increase in Picture3 is significant on a 5% significance level, but it's not as significant as TVsetTV2. We also notice the bias that comes from each assessor, there ratings varies around the mean with a standard devation of 0.57.

$$
  y_{ikj} = \beta_0 + \beta_{TVset_i} + \beta_{Picture_j} + u_{Assessor_k} + \epsilon_{ijk}
$$
Where the random effect and residuals are normal distributed $u_{Assessor_k} \sim N(0, 0.32)$ and $ \epsilon_{ijk} \sim N(0, 1.38) $

We want to emphasize the "on average" aspect in parameter estimation, since we are working with statistics. In repeated samples, the resulting confidence intervals would contain the true parameter 95% of the time.
```{r}
confint(model)
```


### Test // Diagnostic
```{r}

residplot(model)

```

In the normal QQ-plot we see that the points from our model falls approximatly on the line. Indicating that the residuals of our model is normal distributed. This can also bee seen in the "Fitted vs observed" plot where the lines is centered between the datapoints.  

* The points are nicely spread around the horisontal line in the residual plot. 


* We don't see any systematically deviation from the Q-Q Residuals plot. Meaning the the points are in the quantiles we expect, if the residuals are normal distributed.
* We are looking for outlines in the residuals, but the data looks nice. We don't see any points that are higher than $\sqrtr(2)$. 
* We see that especially two data point have higher influence that the rest, namely 48 and 53, it might make sense to look into this.121
=======


```

In the normal QQ-plot we see that the points lays approximately on the line, this show us that the point falls within the quartiles that we expect from a normal distribution. This validates our model assumption that that the points are normal distributed. This assumption is further validated in the plot for the "residuals vs fitted" where the residual values are nicely spread out over and under the line. This tendency can also bee seen in the "fitted vs observed plot". 

We also get a QQ.plot for the random effect, we get a point for each Assosor falling approximatly on the line, validation the assumption that the random effect are gaussian distributed. 
>>>>>>> e11e7c1a02054e415126f3aad3eed620507eef0c






