---
title: "Exploration"
output: html_document
date: "2025-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Init
```{r}
library(emmeans)
library(multcomp)

library(lme4)
library(lmerTest)

library(ggplot2)
library(dplyr)


```

```{r}
setwd("C:/Users/huste/Documents/Github/businessAnalytics/Correlated Data/Assignment 1")
getwd()

list.files()

TV <- read.table("TV2.txt", header = 1)
TV$Assessor <- as.factor(TV$Assessor)
TV$Repeat   <- as.factor(TV$Repeat)
TV$TVset    <- as.factor(TV$TVset)
TV$Picture    <- as.factor(TV$Picture)


```

## Data
### Means
 - Assessor indicates the assessor (1-8), 
 - TVset the television setting (TV1,TV2,TV3),
 - Picture the picture (1-3), 
 - Repeat the replicate (0,1), 
 - (Target) Coloursaturation denotes the colour saturation score given by the assessor. 

```{r}
TV

vars <- c("TVset", "Assessor", "Repeat", "Picture")

par(mfrow = c(2, 2))

for (v in vars) {
  formula <- as.formula(paste("Coloursaturation ~", v))
  
  boxplot(formula, data = TV,
          main = paste("Colour Saturation by", v),
          xlab = v,
          ylab = "Colour Saturation",
          col = "lightblue")

          means <- tapply(TV$Coloursaturation, TV$TVset, mean)
          points(1:length(means), means, pch = 19, col = "black")
  
  }

par(mfrow = c(1, 1))

```
Black dot is mean
Black bar is the median 
Dotted lines shows the confidence interval
Blue box - Interquartile Range (IQR) (Shows the middle 50% of your data.)

We see a difference in mean and meadian in TVset for TV2. It's interquartile range doesn't overlap TV1 and TV3
Also low variance for Assessor 2 and 7

### Interactions
```{r}
# All pairwise interaction plots among these factors
vars <- c("TVset", "Repeat", "Picture")

op <- par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

pairs <- combn(vars, 2, simplify = FALSE)
for (p in pairs) {
  xfac <- p[[1]]
  trc  <- p[[2]]
  interaction.plot(
    x.factor   = TV[[xfac]],
    trace.factor = TV[[trc]],
    response   = TV$Coloursaturation,
    fun        = mean,
    type       = "b",
    pch        = 19,
    xlab       = xfac,
    ylab       = "Colour Saturation (mean)",
    legend     = TRUE,
    main       = paste("Interaction:", xfac, "×", trc)
  )
} 
```

In TVset:Repeat the lines follows each other meaning that it's unlikly there is an interaction. TVset doesn't seem to have an impact on Repeat and vice versa. The same follows for TVset:Picture. 

For repeat:picture it may look there is some interaction because the patterns doesn't follow each other as closely, but the interval between highest and lowest value are very low, which indicate that our model won't improve much by including it.


## Determining the model
### Interactions

```{r}
model_0 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                TVset:Picture + 
                Repeat:Picture +
                (1 | Assessor),
              
              data = TV
  )


model_1 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                TVset:Picture + 
                (1 | Assessor),
              data = TV
  )

model_2 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                TVset:Repeat + 
                (1 | Assessor),
              data = TV
  )

model_3 <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                (1 | Assessor),
              data = TV
  )

a1 <- anova(model_0, model_1, refit = FALSE, type = 3)
a2 <- anova(model_1, model_2, refit = FALSE, type = 3)
a3 <- anova(model_2, model_3, refit = FALSE, type = 3)

cat("#################### Anova 1 - Repeat:Pictur ####################\n")
print(a1)
cat("p-value =", a1$`Pr(>Chisq)`[2], "\n\n")

cat("#################### Anova 2 - TVset:Picture ####################\n")
print(a2)
cat("p-value =", a2$`Pr(>Chisq)`[2], "\n\n")

cat("#################### Anova 3 - TVset:Repeat####################\n")
print(a3)
cat("p-value =", a3$`Pr(>Chisq)`[2], "\n\n")



```
We fail to reject the null hypothesis $ H_0 : \beta_{Repeat:Picture} = 0 $ on a 95% confidence interval with a p-value of 1. This indicates that there is no statistically significant interaction between Repeat and Picture. This holds for all three interactions on with p-values 0.41, 0.13, 0.30.


### Random effects
```{r}

model_effect <- lmer(Coloursaturation ~ 
                TVset + Repeat + Picture +
                (1 | Assessor),
                data = TV
  )

ranova(model_effect)

```

The p-values is very small indicating systematically difference in how Assessors rate color saturation, so we keep the random effect. 


### Fixed effects
```{r}

anova(model_effect, type = 3)

```
We reject the null hypothesis that $ \mu_{TVset} = 0  $ and $ \beta_{picture} = 0  $ on a significance level of 5%. 
We drop Repeat since it doesn't improve the model.

### Final model
```{r}
model <- lmer(
  Coloursaturation ~ TVset + Picture + (1 | Assessor) - 1,
  data = TV
)

summary(model)

```


### Test // Diagnostic
```{r}




```








