---
title: "L6_Exercises"
output: html_document
date: "2025-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lme4)
library(predictmeans)
```


## Cookie analysis
# Buidling the model
```{r}
cookies <- read.csv("https://www2.compute.dtu.dk/courses/02429/Data/datafiles/cookies.txt")
cookies

# Build a a simple model
cookies$assessor <- factor(cookies$assessor)
cookies$treatm <- factor(cookies$treatm)
model1 <- lm(quality ~ assessor + treatm + assessor:treatm,
  data = cookies)

anova(model1)
summary(model1)
```

# Is it a good model?
```{r}
residplot(model1)
```

* The points are nicely spread around the horisontal line in the residual plot. 
* We don't see any systematically devaiation from the Q-Q Residuals plot. Meaning the the points are in the quantiles we expect, if the residuals are normal distributed.
* We are looking for outliers in the resiudalas, but the data looks nice. We don't see any points that are higher than $\sqrtr(2)$. 
* We see that especially two data point have higher influance that the rest, namely 48 and 53, it might make sense to look into this.121




