---
title: "exam_radon"
output: html_document
date: "2025-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Init
```{r}
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(predictmeans)
library(lattice) # to visualize random effects
```

```{r}
setwd("C:/Users/huste/Documents/Github/businessAnalytics/Correlated Data/Exam")
# setwd("/home/husted42/Documents/github/businessAnalytics/Correlated Data/Exam/")

getwd()

list.files()
df<-read.table("Radon_MN.csv",header = T)
head(df)
```

# I. Exploratory analysis
```{r}
df$county <- as.factor(df$county)
df$floor <- as.factor(df$floor)
unique(df$county)
unique(df$floor)
summary(df$u)
summary(df$y)
```

We see that there are 85 different counties and floor is a binary variable 0 meaning basement and 1 meaning not basement.

```{r}
xtabs(~ county, df)
```
The data is not balanced


```{r}
xtabs(~ county, df)
xtabs(~ floor, df)

barplot(
  xtabs(~ county, df),
  las = 2,                # rotate county labels for readability
  col = "lightblue",
  main = "Number of Observations per County",
  ylab = "Count"
)

barplot(
  xtabs(~ floor, df),
  las = 2,                # rotate county labels for readability
  col = "lightblue",
  main = "Number of Observations per floor",
  ylab = "Count"
)

```

```{r}
hist(df$u.full, breaks = 50, col = "lightgray", main = "Distribution of u.full")
hist(df$y, breaks = 50, col = "lightgray", main = "Distribution of log(radon)")
```


```{r}
ggplot(df, aes(x = u.full, y = y, color = floor)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "u.full vs y by floor")

```

```{r}
cor(df[, c("u.full", "y")])

```



We see that neither of the categorical items are balanced. With 5 counties having many samples: 26 and 70 are above 100 while 2, 19 and 80 are above 50. And 83% of the data having floor = 0.

```{r}
vars <- c("x", "county")

# par(mfrow = c(2, 2))

for (v in vars) {
  formula <- as.formula(paste("y ~", v))
  
  boxplot(formula, data = df,
          main = paste("log(Radon) by", v),
          xlab = v,
          ylab = "log(Radon)",
          col = "lightblue")

          means <- tapply(df$y, df$county, mean)
          points(1:length(means), means, pch = 19, col = "black")
  
  }

par(mfrow = c(1, 1))
```

It's not clear whethere there is a significant  between group variance, we see that we have overlap between the inter quartile ranges for the two groups.  


```{r}


radon_floor_0 <- df %>% 
  filter(floor == 0)

ggplot(radon_floor_0, aes(x = u.full, y = y)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +
  labs(
    x = "u.full",
    y = "y",
    title = "floor = 0 : u.full vs y with Regression Line"
  ) +
  theme_minimal()

model <- lm(y ~ u.full, data = radon_floor_0)
summary(model)

```

```{r}
radon_floor_0 <- df %>% 
  filter(floor == 1)

ggplot(radon_floor_0, aes(x = u.full, y = y)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +
  labs(
    x = "u.full",
    y = "y",
    title = "floor = 1 : u.full vs y with Regression Line"
  ) +
  theme_minimal()

model <- lm(y ~ u.full, data = radon_floor_0)
summary(model)

```

# II. Modeling
## Building the model
Target : y
Uranium : u.full
Floor : x
County : county
### Model 0 - Mean
We start with a simple model and build up from there.
We see that the mean is signifincantly different from 0. We do not accept $ H_0 : \mu = 0 $
$$ y_{i} = \mu + \epsilon_i, \epsilon_i \sim N(0, \sigma^2) $$
```{r}
model_0 <- lm(y ~ 1, df)
anova(model_0)

summary(model_0)
```
### Model 1 - Floor
$$ y_{i} = \mu + \alpha_i + \epsilon_i , \quad \epsilon_i \sim N(0, \sigma^2) $$
```{r}
model_1 <- lm(y ~ floor, df)
anova(model_1)
```

### Model 2 - Uranium
$$ y_{i} = \mu + \alpha_i + \beta x_i+ \epsilon_i , \quad \epsilon_i \sim N(0, \sigma^2) $$
```{r}
model_2 <- lm(y ~ floor + u.full , df)
anova(model_2)
```
### Model 3 - Random effect
$$ y_{i} = \mu + \alpha_i + \beta x_i + \gamma_i + \epsilon_i , \quad \epsilon_i \sim N(0, \sigma^2_{\epsilon}), \quad \gamma_i \sim N(0, \sigma^2_{\gamma}) $$
```{r}
model_3 <- lmer(y ~ floor + u.full + (1|county) , df)
ranova(model_3)
```

```{r}
dotplot(ranef(model_3, condVar=TRUE), strip = FALSE, ylab="County") 
ranef_model_3 <- ranef(model_3, condVar=TRUE)
```



### Model 4 & 5 - Random slopes
We add a random slope to the model to allow the impact of the fixed effects to vary across county.

**Random slope for predictor floor**
By including a random slope for the predictor x, we allow its effect on radon levels to depend on the county. This tests whether counties differ in how strongly basement measurements raise or lower radon levels.

```{r}
model_4 <- lmer(y ~ x + u.full + (x|county) , df)
ranova(model_4)

```

The addition of a random-slope term is not statistically significant (p = 0.062). Therefore, the basement effect appears to be consistent across counties, and the random-intercept model is adequate.

```{r}
ranef_m4 <- ranef(model_4)$county
library(ggplot2)
ggplot(ranef_m4, aes(x = x, y = `(Intercept)`)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(title = "Random Slopes and Intercepts for Counties",
       x = "Slope for floor",
       y = "Intercept") +
#  xlim(-0.6, 0.6) +
#  ylim(-0.4, 0.4) +
  coord_fixed() +   # <-- makes width = height
  theme_minimal()
```
The above graph show the the slopes and intercept for each cluster. We don't see clear patterns in the points exeot one straight line. I won't look into further detail as the effect of the random slope are not significant.

**Random slope for predictor u.full**
We skip this since there is no within-county variation.

### Full model
$$ y_{i} = \mu + \alpha_i + \beta x_i + \gamma_i + \epsilon_i , \quad \epsilon_i \sim N(0, \sigma^2_{\epsilon}), \quad \gamma_i \sim N(0, \sigma^2_{\gamma}) $$

```{r}
model <- lmer(y ~ floor + lu.full + (1|county) , df)
summary(model)
```
$$\hat{\sigma_{\epsilon}} = 0.7584, \quad \hat{\sigma_{\gamma}} = 0.1564  $$
We see that only 4 % of the variance is explained by the random effect.
$$ \frac{0.024}{0.575 + 0.024} = 4 \% $$

```{r}
confint(model)
```
Given our observation we are 95% confident that the population mean $\hat{\mu}$ are between (1.39, 1.54) given floor = 0 and u.full = 0. The slope $ \hat{\gamma} $ are between (0.54, 0.90). And that the effect of floor=1 are between (-08, -0.53)




## Diagnostics
```{r}
residplot(model)
```
```{r}
df[299, ]
df[149, ]

predict(model, newdata = df[299, ])
predict(model, newdata = df[149, ])
predict(model, newdata = df[299, ], re.form = NA)
predict(model, newdata = df[149, ], re.form = NA)

```


## Comparison
Look at the change to the diagnostic plots as complexity increasses
```{r}
qqnorm(resid(model_0))
qqline(resid(model_0))
qqnorm(resid(model_3))
qqline(resid(model_3))
```

Comparison table
```{r}
AIC(model_0, model_1, model_2, model_3)
BIC(model_0, model_1, model_2, model_3)
logLik(model_0)
logLik(model_1)
logLik(model_2)
logLik(model_3)

```



| Model   | df | logLik    | AIC      | BIC      |
| ------- | -- | --------- | -------- | -------- |
| model_0 | 2  | –1157.740 | 2319.479 | 2329.126 |
| model_1 | 3  | –1123.512 | 2253.025 | 2267.495 |
| model_2 | 4  | –1064.810 | 2137.621 | 2156.914 |
| model_3 | 5  | –1067.093 | 2144.186 | 2168.303 |

We want AIC and BIC to be as small as possible, but it seems that both AIC and BIC is larger for model 2 than 3.
$$ \text{AIC} = -2 \log L(\hat{\theta}) + 2p, $$
$$ \text{BIC} = -2 \log L(\hat{\theta}) + (\log n)\, p $$

This is partially due to the increase in the complexity penalty since p increases by 1, but we also see an increase in log likelyhood. 

Even though the variance of the residuals decrease from $0.772^2 = 0.596$ in model_2 to $0.758^2 = 0.575$ in model_3, we get an additional variance from the random effect of $0.156^2 = 0.024$ meaning that the overall variance increases hich decreasses the log-likeyhood function. 

```{r}
summary(model_2)
summary(model_3)
```



# III. Results
## III a) Present a plot analogous to the one in Figure 1, with the estimated response means per county

```{r}
model <- lmer(y ~ floor + u.full + (1|county) , df)
df_pred <- df
df_pred_before_random <- df
df_pred$y <- predict(model, newdata = df)
df_pred_before_random$y <- predict(model_2, newdata = df)



invisible({
  attach(df)
  (J <- length(unique(county.name)))
  (ybarbar = mean(y))
  
  (sample.size <- as.vector(table(county)))
  (sample.size.jittered <- sample.size*exp(runif (J, -.1, .1)))
  (cty.mns = tapply(y,county,mean))
  (cty.vars = tapply(y,county,var))
  (cty.sds = mean(sqrt(cty.vars[!is.na(cty.vars)]))/sqrt(sample.size))
  (cty.sds.sep = sqrt(tapply(y,county,var)/sample.size))
  
  par(mfrow=c(1,1))
  plot (sample.size.jittered, cty.mns, cex.lab=.9, cex.axis=1,
        xlab="sample size in county j",
        ylab="Mean log radon in county j",
        pch=20, log="x", cex=.3, mgp=c(1.5,.5,0),
        ylim=c(0,3.2), yaxt="n", xaxt="n")
  axis (1, c(1,3,10,30,100), cex.axis=.9, mgp=c(1.5,.5,0))
  axis (2, seq(0,3), cex.axis=.9, mgp=c(1.5,.5,0))
  for (j in 1:J){
    lines (rep(sample.size.jittered[j],2),
           cty.mns[j] + c(-1,1)*cty.sds[j], lwd=.5)
    #         cty.mns[j] + c(-1,1)*mean(cty.sds[!is.na(cty.sds)]), lwd=.5)
  }
  abline(h=mean(cty.mns),lwd=.5 )
  title("Observed mean response values per county (data)",cex.main=.9, line=1)
  
  cty.mns[cty.mns==max(cty.mns)]
  cty.mns[cty.mns==min(cty.mns)]
  
  points(sample.size.jittered[36],cty.mns[36],cex=2) #max
  points(sample.size.jittered[37],cty.mns[37],cex=2) #min
  detach(df)
  
  attach(df_pred)
  (J <- length(unique(county.name)))
  (ybarbar = mean(y))
  
  (sample.size <- as.vector(table(county)))
  (sample.size.jittered <- sample.size*exp(runif (J, -.1, .1)))
  (cty.mns = tapply(y,county,mean))
  (cty.vars = tapply(y,county,var))
  (cty.sds = mean(sqrt(cty.vars[!is.na(cty.vars)]))/sqrt(sample.size))
  (cty.sds.sep = sqrt(tapply(y,county,var)/sample.size))
  
  par(mfrow=c(1,1))
  plot (sample.size.jittered, cty.mns, cex.lab=.9, cex.axis=1,
        xlab="sample size in county j",
        ylab="Mean log radon in county j",
        pch=20, log="x", cex=.3, mgp=c(1.5,.5,0),
        ylim=c(0,3.2), yaxt="n", xaxt="n")
  axis (1, c(1,3,10,30,100), cex.axis=.9, mgp=c(1.5,.5,0))
  axis (2, seq(0,3), cex.axis=.9, mgp=c(1.5,.5,0))
  for (j in 1:J){
    lines (rep(sample.size.jittered[j],2),
           cty.mns[j] + c(-1,1)*cty.sds[j], lwd=.5)
    #         cty.mns[j] + c(-1,1)*mean(cty.sds[!is.na(cty.sds)]), lwd=.5)
  }
  abline(h=mean(cty.mns),lwd=.5 )
  title("Observed mean response values per county full_model",cex.main=.9, line=1)
  
  cty.mns[cty.mns==max(cty.mns)]
  cty.mns[cty.mns==min(cty.mns)]
  
  points(sample.size.jittered[36],cty.mns[36],cex=2) #max
  points(sample.size.jittered[37],cty.mns[37],cex=2) #min
  
  # print(cty.sds)
  
  detach(df_pred)
  
  attach(df_pred_before_random)
  (J <- length(unique(county.name)))
  (ybarbar = mean(y))
  
  (sample.size <- as.vector(table(county)))
  (sample.size.jittered <- sample.size*exp(runif (J, -.1, .1)))
  (cty.mns = tapply(y,county,mean))
  (cty.vars = tapply(y,county,var))
  (cty.sds = mean(sqrt(cty.vars[!is.na(cty.vars)]))/sqrt(sample.size))
  (cty.sds.sep = sqrt(tapply(y,county,var)/sample.size))
  
  par(mfrow=c(1,1))
  plot (sample.size.jittered, cty.mns, cex.lab=.9, cex.axis=1,
        xlab="sample size in county j",
        ylab="Mean log radon in county j",
        pch=20, log="x", cex=.3, mgp=c(1.5,.5,0),
        ylim=c(0,3.2), yaxt="n", xaxt="n")
  axis (1, c(1,3,10,30,100), cex.axis=.9, mgp=c(1.5,.5,0))
  axis (2, seq(0,3), cex.axis=.9, mgp=c(1.5,.5,0))
  for (j in 1:J){
    lines (rep(sample.size.jittered[j],2),
           cty.mns[j] + c(-1,1)*cty.sds[j], lwd=.5)
    #         cty.mns[j] + c(-1,1)*mean(cty.sds[!is.na(cty.sds)]), lwd=.5)
  }
  abline(h=mean(cty.mns),lwd=.5 )
  title("Observed mean response values per county Model 1",cex.main=.9, line=1)
  
  cty.mns[cty.mns==max(cty.mns)]
  cty.mns[cty.mns==min(cty.mns)]
  
  points(sample.size.jittered[36],cty.mns[36],cex=2) #max
  points(sample.size.jittered[37],cty.mns[37],cex=2) #min
  detach(df_pred_before_random)
})

```

There should be at max two differnt outcomes county
```{r}
df_pred %>%
  group_by(county) %>%
  summarise(n_unique_y = n_distinct(y)) %>% head(10)
```
```{r}
df_pred %>%
  group_by(county) %>%
  summarise(
    y_values = list(sort(unique(y))),
    diff_y = diff(sort(unique(y)))
  ) %>% head(10)
```


```{r}
overall_mean <- mean(df_pred$y)

df_pred %>%
  group_by(county) %>%
  summarise(
    county_mean = mean(y),
    diff_from_overall = county_mean - overall_mean
  ) %>%
  arrange(desc(diff_from_overall)) %>%
  head(10)
```


## III b) Present a plot of the county level of uranium in the horizontal axis, and the estimated regression intercepts with CI, in the vertical axis. Additionally, one with the estimated regression slopes with CI.


```{r}
model <- lmer(y ~ floor + u.full + (1|county) , df)

library(lme4)
library(ggplot2)

# Random intercepts with conditional variance
re  <- ranef(model, condVar=TRUE)
u   <- attr(re$county, "postVar")   # 1x1xN array of variances

# Random-intercept estimates
intercepts <- re$county[, "(Intercept)"]

# Std. errors from conditional modes
se <- sqrt(sapply(1:dim(u)[3], function(i) u[,,i]))

# Build dataframe
df_plot <- data.frame(
  county    = rownames(re$county),
  intercept = intercepts,
  lower     = intercepts - 1.96 * se,
  upper     = intercepts + 1.96 * se
)

u_by_county <- aggregate(u.full ~ county, data=df, FUN=unique)
names(u_by_county)[2] <- "u_full_county"

df_plot$u_full <- u_by_county$u_full_county[
  match(df_plot$county, u_by_county$county)
]

ggplot(df_plot, aes(x = u_full, y = intercept)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0) +
  xlab("County-level uranium (u.full)") +
  ylab("Random intercept with 95% CI") +
  ylim(-0.5, 0.5) +
  theme_bw()


```

```{r}
# Posterior variance for each county's random intercept
post_var <- sapply(1:dim(u)[3], function(i) u[,,i])

# Build a dataframe with county, variance, SE, and intercept
var_by_county <- data.frame(
  county    = rownames(re$county),
  intercept = re$county[, "(Intercept)"],
  var       = post_var,
  se        = sqrt(post_var)
)

# Counties with the *lowest* conditional variance
var_by_county_lowest <- var_by_county %>% 
  arrange(var)

# View the top few with lowest variance
head(var_by_county_lowest, 10)

var_by_county_highest <- var_by_county %>% 
  arrange(desc(var))

# View the top counties with the largest variance
head(var_by_county_highest, 10)
```


## III c) Present a plot of the county level of uranium in the horizontal axis, and the estimated regression intercepts with CI, in the vertical axis. Additionally, one with the estimated regression slopes with CI.

```{r}
# 6 in sample
# 6 out of sample

new_data <- data.frame(
  county = c(
            35,35,
            43,43,
            65,65,
            100,100,
            101,101,
            102,102
            ),
  floor  = c(
            0,1,
            0,1,
            0,1,
            0,1,
            0,1,
            0,1
            ),
  u.full = c(
            -0.8818289, -0.8818289,
            0.013764829,0.013764829, 
            0.52802487,0.52802487, 
            -1,-1, 
            0,0,
            0.7,0.7
            )
)


new_data$floor  <- factor(new_data$floor)
new_data$county <- factor(new_data$county)
new_data$pred_conditional <- predict(model_3, newdata = new_data, re.form = NA)

new_data

```

```{r}
# install.packages("merTools")
library(merTools)

new_data$floor  <- factor(new_data$floor)
new_data$county <- factor(new_data$county)

pred_ci <- predictInterval(
  model,
  newdata = new_data,
  which   = "full",
  level   = 0.95,
  n.sims  = 1000,
  include.resid.var = TRUE
)

new_data <- cbind(new_data, pred_ci)

```


```{r}
library(lme4)

## Design matrix for the fixed-effects part of the model
## This must match the fixed-effects formula: ~ floor + u.full
X <- model.matrix(~ floor + u.full, data = new_data)

## Fixed-effect estimates and their covariance matrix
beta_hat <- fixef(model)
V_beta   <- vcov(model)

## Standard errors of the linear predictor X * beta
pred_se <- sqrt(diag(X %*% V_beta %*% t(X)))

## 95% confidence intervals
new_data$pred_se  <- pred_se
new_data$lower_ci <- new_data$pred_conditional - 1.96 * pred_se
new_data$upper_ci <- new_data$pred_conditional + 1.96 * pred_se

library(ggplot2)

ggplot(new_data, aes(x = u.full, y = pred_conditional, colour = county)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0) +
  xlab("u.full") +
  ylab("Predicted y (95% CI, fixed effects only)") +
  theme_bw()


```

