---
title: "Part 2"
output: html_document
date: "2025-12-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Init
```{r}
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(predictmeans)
library(lattice) # to visualize random effects
library(nlme) # Rondom thingy
library(tidyr)
library(tidyverse)

```

```{r}
# setwd("C:/Users/huste/Documents/Github/businessAnalytics/Correlated Data/Exam")
setwd("/home/husted42/Documents/github/businessAnalytics/Correlated Data/Exam/")

getwd()

list.files()
rats_short <- read.table("Rats_Box.txt", header = 1, sep = ",")
head(rats_short)



```


# i) Prepare for analysis
We want to measure the gain of each rat, so our response variable should only be the gain y_0,...,y_4. But we need to use y_0 as an adjusting covariate in the model, ie. we should keep it as a feature. Practically this means that we filter out y_0, convert the data to long format, then join y_0 back again.
 

```{r}
# Pivot to long format
rats_long <- rats_short %>%
  pivot_longer(cols = starts_with("y"),
               names_to = "Week",
               values_to = "Value") %>%
  mutate(
    Week = as.numeric(substr(Week, 2, 2))
  ) %>%
  arrange(Rat, Week)

# Extract the first week
rats_y0 <- rats_long %>%
  dplyr::filter(Week == 0) %>%
  dplyr::select(Rat, Trt, y0 = Value)

# Filter out the first week
rats_gain <- rats_long %>%
  filter(Week != 0)

# Left join the y0 values
rats_long_gain <- rats_gain %>%
  left_join(rats_y0, by = c("Rat", "Trt"))

# NOTE : We should include y_0 in the report
rats <- rats_long_gain
rats$WeekQ <- rats$Week
rats$Week <- paste0("y", rats$Week)
rats$Rat <- as.factor(rats$Rat)
rats$Trt <- as.factor(rats$Trt)
rats$Week <- as.factor(rats$Week)
colnames(rats)[colnames(rats) == 'y0'] <- 'InitWheight'
colnames(rats)[colnames(rats) == 'Value'] <- 'Gain'
rats <- rats %>%
  dplyr::select(Rat, Trt, Week, WeekQ, InitWheight, Gain)

rats$InitWheight <- as.numeric(rats$InitWheight)
rats$Gain <- as.numeric(rats$Gain)

head(rats)


```


```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Rat)) +
  geom_line(alpha = 0.3) +
  geom_point() +
  labs(title = "Weight Gain of Rats Over Weeks by Treatment",
       x = "Week",
       y = "Weight Gain") +
  theme_minimal()
```
```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Trt)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 2.5) +
  labs(title = "Mean Weight Gain of Rats Over Weeks by Treatment",
       x = "Week",
       y = "Mean Weight Gain") +
  theme_minimal()
```

# ii)  Do an exploratory analysis of the data, numerical and graphical

```{r}
#str(rats)
summary(rats$InitWheight)
summary(rats$Gain)
```

When just looking at the distribution of gain, without accounting for treatment, the data look appproximatly nomrlal distributed.
```{r}
ggplot(rats, aes(Gain)) +
  geom_histogram(bins = 15, color="black", fill="grey70") +
  theme_minimal() +
  labs(title = "Distribution of Weight Gain")
```

The within-treatment variability seems to be larger for treatments than the control group. The the difference in mean are smaller between control group and treatment-2, with a slight increase in quantiles ranges and variability. It not clear if the effect of treatment-2 is significant from the plot. However it looks like there is a noticeably difference between treatment-3 and the control group, which indicates that we the effect of treatment-3 is significant. But we cannot draw any conclusions.
```{r}
ggplot(rats, aes(x = Trt, y = Gain, fill = Trt)) +
  geom_boxplot() +
  labs(title = "Gain by Treatment") +
  theme_minimal()
```

By this plot it doesn't look like time alone has a significant effect on the increase in weights. Other than there seems to be an increase in the variability between week 1-2 and 3-4.
```{r}
ggplot(rats, aes(x = Week, y = Gain)) +
  geom_boxplot() +
  labs(title = "Gain by Week", x = "Week") +
  theme_minimal()
```

We check to see if there are correlations between the initial weights and gain. But with a correlation of 0.02, there is no clear evidence of this as is also seen in the figure.
```{r}
cor(rats$InitWheight, rats$Gain)
```
```{r}
rats %>%
  dplyr::group_by(Rat, Trt) %>%
  dplyr::summarise(mean_gain = mean(Gain), InitWheight = first(InitWheight), .groups = "drop") %>%
  ggplot(aes(x = InitWheight, y = mean_gain, color = Trt)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE) +
    theme_minimal() +
    labs(title = "Initial Weight vs. Gain")
```
TODO : Interaction plot for group and time

```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Trt)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(
    x = "Week",
    y = "Mean Gain",
    color = "Trt",
    title = "Interaction Plot: Treatment by Week"
  ) +
  theme_minimal()
```


# iii) Formulate a mixed model
We will build a mixed model from the button up, starting with the mean and a random effect for the rats
## Model 0 - Random effect
```{r}
mll_0 <- lmer(Gain ~ 1 + (1|Rat), data = rats)
ranova(mll_0)
```
## Model 0.1 - InitWheight
```{r}
mll_0_1 <- lmer(Gain ~ InitWheight + (1|Rat), data = rats)
anova(mll_0_1)
```

## Model 1 - Adding treatment
```{r}
mll_1 <- lmer(Gain ~ Trt + (1|Rat), data = rats)
anova(mll_1)
```
## Model 2 - Adding fixed Week
```{r}
mll_2 <- lmer(Gain ~ Trt + Week + (1|Rat), data = rats)
anova(mll_2)
```
## Model 3 - Adding interaction
```{r}
mll_3 <- lmer(Gain ~ Trt + factor(Week) + Trt:Week + (1|Rat), data = rats)
anova(mll_3)
```

### Diagnostic
```{r}
dotplot(ranef(mll_3, condVar=TRUE), strip = FALSE, ylab="Rat") 
ranef_model_3 <- ranef(mll_3, condVar=TRUE)
```

```{r}
residplot(mll_3)
```

## Model 4 - Numerical Weeks
```{r}
mll_4 <- lmer(Gain ~ Trt + WeekQ + Trt:WeekQ + (1|Rat), data = rats)
anova(mll_4)
ranova(mll_4)
```
```{r}
residplot(mll_4)
```

```{r}
anova(mll_3, mll_4)
```


### Diagnostic
```{r}
residplot(mll_4)
```


## Comapare
```{r}
anova(mll_3, mll_4)
```

# Random effect thinghy
## Create models
```{r}
## 0) Make sure data is right
rats_df <- as.data.frame(rats)
rats_df$Rat <- factor(as.character(rats_df$Rat))
rats_df$Trt <- factor(as.character(rats_df$Trt))
# Week is fine as factor for the fixed effect
# Ensure WeekQ is numeric
rats_df$WeekQ <- as.numeric(rats_df$WeekQ)


## 1) Compound Symmetry (corCompSymm)
M1 <- lme(Gain ~ WeekQ + Trt + Trt:WeekQ, random = ~ 1 | Rat,
          correlation = corCompSymm(form = ~ WeekQ | Rat),
          data = rats_df)

## 2) AR(1) (corAR1)
M2 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corAR1(form = ~ WeekQ | Rat),
  data = rats_df
)

## 3) ARMA(1,1) (corARMA)
M3 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corARMA(p = 1, q = 1, form = ~ WeekQ | Rat),
  data = rats_df
)


## 4) Continuous AR(1) (corCAR1)
M4 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corCAR1(form = ~ WeekQ | Rat),
  data = rats_df
)


## 5) Exponential spatial (corExp)
M5 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corExp(form = ~ WeekQ | Rat),
  data = rats_df
)


## 6) Gaussian spatial (corGaus)
M6 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corGaus(form = ~ WeekQ | Rat),
  data = rats_df
)


## 7) Linear spatial (corLin)
M7 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corLin(form = ~ WeekQ | Rat),
  data = rats_df
)


## 8) Rational quadratic spatial (corRatio)
M8 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corRatio(form = ~ WeekQ | Rat),
  data = rats_df
)


## 9) Spherical spatial (corSpher)
M9 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corSpher(form = ~ WeekQ | Rat),
  data = rats_df
)


## 10) Unstructured correlation (corSymm)
M10 <- lme(
  Gain ~ WeekQ + Trt + Trt:WeekQ,
  random = ~ 1 | Rat,
  correlation = corSymm(form = ~ 1 | Rat),
  data = rats_df
)


```

## Compare
```{r}
models <- list(
  M1 = M1, 
  M2 = M2, 
  M3 = M3, 
  M4 = M4, 
  M5 = M5, 
  M6 = M6, 
  M7 = M7, 
  M8 = M8, 
  M9 = M9, 
  M10 = M10
)

models_names <- c(
  "corCompSymm",
  "corAR1",
  "corARMA",
  "corCAR1",
  "corExp",
  "corGaus",
  "corLin",
  "corRatio",
  "corSpher",
  "corSymm"
)

model_comp <- data.frame(
  Model = names(models),
  Name = models_names,
  AIC  = round(sapply(models, AIC), 2),
  BIC  = round(sapply(models, BIC), 2),
  logLik = round(sapply(models, function(x) logLik(x)[1]), 2)
) %>%
  arrange(AIC)

model_comp

```



## Viogram plots
```{r fig.show='hold'}
models <- list(
  M1 = M1, M2 = M2, M3 = M3, M4 = M4, M5 = M5,
  M6 = M6, M7 = M7, M8 = M8, M9 = M9, M10 = M10
)

for (name in names(models)) {
  vg <- suppressWarnings(
    Variogram(models[[name]], form = ~ WeekQ | Rat, data = rats_df)
  )
  
  print(
    plot(
      vg,
      cex = 1,
      pch = 16,
      ylim = c(0, 2),
      xlab = paste("Distance:", name),
      main = name
    )
  )
}
```

```{r}
par(mfrow=c(1,3))

## AR1 vs CAR1
plot(predict(M2, Trt=1), predict(M4, Trt=1), pch=20,
     main = "M2 (AR1) vs M4 (CAR1)",
     xlab = "Predictions from M2 (AR1)",
     ylab = "Predictions from M4 (CAR1)")
abline(0,1,col="red")

## AR1 vs Exponential
plot(predict(M2, Trt=1), predict(M5, Trt=1), pch=20,
     main = "M2 (AR1) vs M5 (Exponential)",
     xlab = "Predictions from M2 (AR1)",
     ylab = "Predictions from M5 (Exponential)")
abline(0,1,col="red")

## Exponential vs CAR1
plot(predict(M5, Trt=1), predict(M4, Trt=1), pch=20,
     main = "M5 (Exponential) vs M4 (CAR1)",
     xlab = "Predictions from M5 (Exponential)",
     ylab = "Predictions from M4 (CAR1)")
abline(0,1,col="red")


```

```{r}
residplot(M2)

```

residuals(object, level = Q,
          type = c("response", "pearson", "normalized"), asList = FALSE, ...)z





```{r}
summary(mll_4)
# summary(M2)

confint(mll_4)
intervals(M2, which = 'fixed')
```
Make a table showing the fixed effects estimates, standard errors, t-values, and p-values. 

```{r}
summary(M2)

library(emmeans) #estimated response values
emmeans(M3, "Trt", by="WeekQ", data=rats)

```


