---
title: "Part 2"
output: html_document
date: "2025-12-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Init
```{r}
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(predictmeans)
library(lattice) # to visualize random effects
```

```{r}
setwd("C:/Users/huste/Documents/Github/businessAnalytics/Correlated Data/Exam")
# setwd("/home/husted42/Documents/github/businessAnalytics/Correlated Data/Exam/")

getwd()

list.files()
rats_short <- read.table("Rats_Box.txt", header = 1, sep = ",")
head(rats_short)



```


# i) Prepare for analysis
We want to measure the gain of each rat, so our response variable should only be the gain y_0,...,y_4. But we need to use y_0 as an adjusting covariate in the model, ie. we should keep it as a feature. Practically this means that we filter out y_0, convert the data to long format, then join y_0 back again.
 

```{r}
# Pivot to long format
rats_long <- rats_short %>%
  pivot_longer(cols = starts_with("y"),
               names_to = "Week",
               values_to = "Value") %>%
  mutate(
    Week = as.numeric(substr(Week, 2, 2))
  ) %>%
  arrange(Rat, Week)

# Extract the first week
rats_y0 <- rats_long %>%
  filter(Week == 0) %>%
  select(Rat, Trt, y0 = Value)

# Filter out the first week
rats_gain <- rats_long %>%
  filter(Week != 0)

# Left join the y0 values
rats_long_gain <- rats_gain %>%
  left_join(rats_y0, by = c("Rat", "Trt"))

# NOTE : We should include y_0 in the report
rats <- rats_long_gain
rats$WeekQ <- rats$Week
rats$Week <- paste0("y", rats$Week)
rats$Rat <- as.factor(rats$Rat)
rats$Trt <- as.factor(rats$Trt)
rats$Week <- as.factor(rats$Week)
colnames(rats)[colnames(rats) == 'y0'] <- 'InitWheight'
colnames(rats)[colnames(rats) == 'Value'] <- 'Gain'
rats <- rats %>%
  select(Rat, Trt, Week, WeekQ, InitWheight, Gain)

rats$InitWheight <- as.numeric(rats$InitWheight)
rats$Gain <- as.numeric(rats$Gain)

head(rats)


```


```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Rat)) +
  geom_line(alpha = 0.3) +
  geom_point() +
  labs(title = "Weight Gain of Rats Over Weeks by Treatment",
       x = "Week",
       y = "Weight Gain") +
  theme_minimal()
```
```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Trt)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 2.5) +
  labs(title = "Mean Weight Gain of Rats Over Weeks by Treatment",
       x = "Week",
       y = "Mean Weight Gain") +
  theme_minimal()
```

# ii)  Do an exploratory analysis of the data, numerical and graphical

```{r}
#str(rats)
summary(rats$InitWheight)
summary(rats$Gain)
```

When just looking at the distribution of gain, without accounting for treatment, the data look appproximatly nomrlal distributed.
```{r}
ggplot(rats, aes(Gain)) +
  geom_histogram(bins = 15, color="black", fill="grey70") +
  theme_minimal() +
  labs(title = "Distribution of Weight Gain")
```

The within-treatment variability seems to be larger for treatments than the control group. The the difference in mean are smaller between control group and treatment-2, with a slight increase in quantiles ranges and variability. It not clear if the effect of treatment-2 is significant from the plot. However it looks like there is a noticeably difference between treatment-3 and the control group, which indicates that we the effect of treatment-3 is significant. But we cannot draw any conclusions.
```{r}
ggplot(rats, aes(x = Trt, y = Gain, fill = Trt)) +
  geom_boxplot() +
  labs(title = "Gain by Treatment") +
  theme_minimal()
```

By this plot it doesn't look like time alone has a significant effect on the increase in weights. Other than there seems to be an increase in the variability between week 1-2 and 3-4.
```{r}
ggplot(rats, aes(x = Week, y = Gain)) +
  geom_boxplot() +
  labs(title = "Gain by Week", x = "Week") +
  theme_minimal()
```

We check to see if there are correlations between the initial weights and gain. But with a correlation of 0.02, there is no clear evidence of this as is also seen in the figure.
```{r}
cor(rats$InitWheight, rats$Gain)
```
```{r}
rats %>%
  dplyr::group_by(Rat, Trt) %>%
  dplyr::summarise(mean_gain = mean(Gain), InitWheight = first(InitWheight), .groups = "drop") %>%
  ggplot(aes(x = InitWheight, y = mean_gain, color = Trt)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE) +
    theme_minimal() +
    labs(title = "Initial Weight vs. Gain")
```
TODO : Interaction plot for group and time

```{r}
ggplot(rats, aes(x = Week, y = Gain, color = Trt, group = Trt)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(
    x = "Week",
    y = "Mean Gain",
    color = "Trt",
    title = "Interaction Plot: Treatment by Week"
  ) +
  theme_minimal()
```


# iii) Formulate a mixed model
We will build a mixed model from the button up, starting with the mean and a random effect for the rats
## Model 0 - Random effect
```{r}
mll_0 <- lmer(Gain ~ 1 + (1|Rat), data = rats)
ranova(mll_0)
```
## Model 1 - Adding treatment
```{r}
mll_1 <- lmer(Gain ~ Trt + (1|Rat), data = rats)
summary(mll_1)
anova(mll_1)
ranova(mll_1)
```
## Model 2 - Adding fixed Week
```{r}
mll_2 <- lmer(Gain ~ Trt + Week + (1|Rat), data = rats)
anova(mll_2)
```
## Model 3 - Adding interaction
```{r}
mll_3 <- lmer(Gain ~ Trt + Week + Trt:Week + (1|Rat), data = rats)
anova(mll_3)
```

### Diagnostic
```{r}
dotplot(ranef(mll_3, condVar=TRUE), strip = FALSE, ylab="Rat") 
ranef_model_3 <- ranef(mll_3, condVar=TRUE)
```

```{r}
residplot(mll_3)
```

## Model 4 - Numerical Weeks
```{r}
mll_4 <- lmer(Gain ~ Trt + WeekQ + Trt:Week + (1|Rat), data = rats)
anova(mll_4)
ranova(mll_4)
```
```{r}
residplot(mll_4)
```

