---
title: "WLS"
output: html_document
date: "2026-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
### Read training data
#! Perhaps you need to set the working directory!?
#setwd("/home/pbac/g/course02417/2025/assignment1")
D <- read.csv("DST_BIL54.csv")
str(D)

# See the help
?strftime
D$time <- as.POSIXct(paste0(D$time,"-01"), "%Y-%m-%d", tz="UTC")
D$time
class(D$time)

## Year to month for each of them
D$year <- 1900 + as.POSIXlt(D$time)$year + as.POSIXlt(D$time)$mon / 12

## Make the output variable a floating point (i.e.\ decimal number)
D$total <- as.numeric(D$total) / 1E6

## Divide intro train and test set
teststart <- as.POSIXct("2024-01-01", tz="UTC")
Dtrain <- D[D$time < teststart, ]
Dtest <- D[D$time >= teststart, ]

### convert time to a variable x
D$x <- 1900 + as.POSIXlt(D$time)$year + as.POSIXlt(D$time)$mon / 12
D$x <- as.numeric(format(D$time, "%Y")) +
          (as.numeric(format(D$time, "%m")) - 1) / 12

# Only keep x and total in d
D <- D[, c("x", "total", "year")]
D$x <- D$x - 2018

Dsmall <- D[, c("x", "total")]

Dtrain <- Dsmall[Dsmall$x < (2024 - 2018), ]
Dtest  <- Dsmall[Dsmall$x >= (2024 - 2018), ]

```



# WLS

We setup the weights st. we have $ w_n = \lambda^0, w_{n-1} = \lambda^1, w_{n-2}  = \lambda²... , w_1 = \lambda^n $. We can use these weight to make our covariance matrix $ \Sigma $. We are not interested in modelling the correlation between timepoints, meaning that we will only have values in the diagonal. We want to model it st. observations with low weight allows for a high variance. As in eq. 3.96 we pick $ \Sigma = \text{Diag[} 1 / \lambda^{n-1}, ... , 1 / \lambda, 1  $  

```{r}
n <- length(Dtrain$x)
n
lambda = 0.9
weights <- lambda^((n-1):0)
# plot the weights:
barplot(weights, names=1:n)


SIGMA <- diag(n)
diag(SIGMA) <- 1/weights
# print lower right corner to check:
print(SIGMA[(n-6):n, (n-6):n]) # looks fine :-)

```
### 
```{r}
sum_wls <- sum(weights)
sum_wls

```

# Estimate parameters
```{r}
# GLS / generalized least squares estimator:
# theta_hat = (t(X) %*% Sigma^{-1} %*% X)^{-1} %*% t(X) %*% Sigma^{-1} %*% Y
gls_theta_hat <- function(X, Sigma, Y, tol = 1e-10) {
  X <- as.matrix(X)
  Sigma <- as.matrix(Sigma)
  Y <- as.matrix(Y)
  if (ncol(Y) != 1) stop("Y must be a vector or an n x 1 matrix.")
  if (nrow(X) != nrow(Sigma) || nrow(Sigma) != nrow(Y)) {
    stop("Dimensions must match: nrow(X) = nrow(Sigma) = length(Y).")
  }

  # Compute Sigma^{-1} X and Sigma^{-1} Y without forming Sigma^{-1}
  SinvX <- solve(Sigma, X)
  SinvY <- solve(Sigma, Y)

  A <- t(X) %*% SinvX         # = t(X) Sigma^{-1} X
  b <- t(X) %*% SinvY         # = t(X) Sigma^{-1} Y

  # Check full rank of A
  r <- qr(A, tol = tol)$rank
  if (r < ncol(X)) {
    stop("t(X) %*% Sigma^{-1} %*% X is not full rank (or nearly singular).")
  }

  # Solve A * theta = b
  theta_hat <- solve(A, b)
  drop(theta_hat)  # return as numeric vector
}

# Example usage:
X <- cbind(1, Dtrain$x)        # include intercept if needed
theta_hat <- gls_theta_hat(X, SIGMA, Dtrain$total)
print(theta_hat)
```

```{r}
w <- weights  # from your code above

fit_wls <- lm(total ~ x, data = Dtrain, weights = w)
summary(fit_wls)
```

```{r}
X <- model.matrix(total ~ x, data = Dtrain)
Y <- Dtrain$total
theta_hat_ols <- solve(t(X) %*% X) %*% t(X) %*% Y
theta_hat_wls <- theta_hat

plot(Dtrain$x, Dtrain$total,
     type = "p",
     xlab = "Year",
     ylab = "Total (millions)",
     main = "Total Vehicles Over Time with Linear Fit")

abline(a = theta_hat_wls[1], b = theta_hat_wls[2], col = "red", lwd = 2)
abline(a = theta_hat_ols[1], b = theta_hat_ols[2], col = "blue", lwd = 2)
```


# Forecast
```{r}
# Predictions
pred_wls <- predict(lm_fit_wls, newdata = Dtest, interval = "prediction")
pred_ols <- predict(lm_fit_ols, newdata = Dtest, interval = "prediction")

# Make sure x is ordered for clean lines
ord <- order(Dtest$x)

x_vals <- Dtest$x[ord]
y_vals <- Dtest$total[ord]

pred_wls <- pred_wls[ord, ]
pred_ols <- pred_ols[ord, ]

# Set y-limits to include both models' intervals
ylim_all <- range(
  pred_wls[, c("lwr","upr")],
  pred_ols[, c("lwr","upr")]
)

# Base plot (WLS)
plot(x_vals, pred_wls[, "fit"], type = "l", lwd = 2,
     ylim = ylim_all,
     xlab = "x", ylab = "total",
     main = "WLS vs OLS Forecast with Prediction Intervals")

# WLS intervals
lines(x_vals, pred_wls[, "lwr"], lty = 2, lwd = 2)
lines(x_vals, pred_wls[, "upr"], lty = 2, lwd = 2)

# OLS fit
lines(x_vals, pred_ols[, "fit"], lwd = 2, col = 2)

# OLS intervals
lines(x_vals, pred_ols[, "lwr"], lty = 2, col = 2)
lines(x_vals, pred_ols[, "upr"], lty = 2, col = 2)

# Test data points
points(x_vals, y_vals, pch = 16)

# Legend
legend("topleft",
       legend = c("WLS fit", "WLS PI",
                  "OLS fit", "OLS PI",
                  "Test data"),
       col = c(1, 1, 2, 2, 1),
       lty = c(1, 2, 1, 2, NA),
       lwd = c(2, 2, 2, 2, NA),
       pch = c(NA, NA, NA, NA, 16),
       bty = "n")
```

